Taking Dyads Seriously

### 2018-10-02 paper/1_intro.tex
CD edits

We begin with a discussion of these dependencies and an introduction to the AME model. Next, we conduct a simulation study to show how the AME approach can recover unbiased and well-calibrated regression coefficients in the context of dyadic data. Last, to highlight the utility of this approach, we apply the AME model to three recent studies in the international relations (IR) literature (published in this journal). Our comparison reveals that by accounting for observational dependence, AME produces more precise estimates and better-calibrated confidence intervals for key variables. Consequently, AME generates results that, at times, differ from those found in the original study as well as from the broader literature. Moreover, we demonstrate that the AME latent factor approach offers substantive insights that are often occluded by ignoring the interdependencies found in the relational data widely employed in IR. Finally, we show that for each replication our network-based approach provides substantially more accurate out-of-sample predictions than the models used in the original studies.

Our AME approach advances statistical analysis of dyadic data by accounting for observational dependence while allowing scholars to test the substantive effect of variables of interest. Thus, the AME allows scholars to achieve a two-fold goal: to continue to generate meaningful, substantive insights about political phenomena without abandoning a regression based approach, but at the same time account for the data generating processes behind such events of interest. Most importantly, the AME approach concentrates on the relational aspect of the field of international relations through a statistical framework that is familiar to most scholars.

>>>>

We begin with a discussion of these dependencies and an introduction to the AME model. Next, we conduct a simulation study to show how the AME approach can recover unbiased and well-calibrated regression coefficients in the context of dyadic data. Last, to highlight the utility of this approach, we apply the AME model to three recent studies in the international relations (IR) literature. Our comparison reveals that by accounting for observational dependence, AME produces more precise estimates and better-calibrated confidence intervals for key variables. Consequently, AME generates results that, at times, differ from those found in the original study as well as from the broader literature. Moreover, we demonstrate that the AME latent factor approach offers substantive insights that are often occluded by ignoring the interdependencies found in the relational data widely employed in IR. Finally, we show that for each replication our network-based approach provides substantially more accurate out-of-sample predictions than the models used in the original studies.

The AME approach advances statistical analysis of dyadic data by accounting for observational dependence while allowing scholars to test the substantive effect of variables of interest. Thus, the AME allows scholars to achieve a two-fold goal: to continue to generate meaningful, substantive insights about political phenomena without abandoning a regression based approach, while at the same time accounting for the data generating processes behind such events of interest. Most importantly, the AME approach concentrates on the relational aspect of the field of international relations through a statistical framework that is familiar to most scholars.

### 2018-10-02 paper/2_model.tex
CD edits

Scholars working with dyadic data typically begin by stacking observations associated with each dyad on top of one another. This makes sense if each observation is independent of the others. For example, a conflict initiated from the United States against Japan is assumed to be independent of any conflictual action that Japan may send to the United States. Additionally, every action sent by Japan to others in the system is considered independent even though each of those interactions involves a common sender, i.e, Japan. While most scholars begin with the assumption that each dyadic interaction takes place in isolation of the others, we know this assumption to be false both in theory and in practice. Relational data comes with an explicit structure that generally leads to particular types of dependencies. The importance of accounting for the underlying structure of our data has been a lesson well understood, at least when it comes to time-series cross-sectional data (TSCS) within political science \citep{beck:katz:1995,beck:etal:1998}. As a result, it is now standard practice to take explicit steps to account for the complex data structures that emerge in TSCS applications and the unobserved heterogeneity that they cause.

@@

These types of patterns frequently emerge in IR contexts as well.\footnote{For example, see: \citet{manger:etal:2012, kinne:2013, chyzh:2016}.} For example, a perennial finding in the interstate trade literature emphasizes the role that geography plays in determining trade flows. Geographic proximity in the network context is an example of homophily---a shared attribute between actors that corresponds to a greater likelihood of the event of interest taking place. Alternatively, in the interstate conflict literature, we may find that actors who are each a member of a particular (formal or informal) alliance are likely to act similarly in the conflict network. Specifically, they will tend to initiate conflictual events with actors that their fellow alliance members initiate conflict with, and they will be unlikely to initiate conflict with members of their alliance---an example of stochastic equivalence. In both these examples, we are able to explicitly parameterize the attribute that might explain the emergence of higher order dependence patterns. While sometimes the conditions driving these patterns, such as geography, are easy to identify, at other times it can be difficult to describe exactly why higher order dependence patterns in networks may develop.
%CD to SM ^

@@

% We use a Bayesian probit framework, in which we model a latent variable, $\theta_{ij}$, first using a set of exogenous dyadic ($\bm\beta_{d}^{\top} \mathbf{X}_{ij}$), sender ($\bm\beta_{s}^{\top} \mathbf{X}_{i}$), and receiver covariates ($\bm\beta_{r}^{\top} \mathbf{X}_{j}$).
where $y_{ij,t}$ captures the interaction between actor $i$ (the sender) and $j$ (the receiver) at time $t$. We model a latent variable, $\theta_{ij}$, first using a set of exogenous dyadic ($\bm\beta_{d}^{\top} \mathbf{X}_{ij}$), sender ($\bm\beta_{s}^{\top} \mathbf{X}_{i}$), and receiver covariates ($\bm\beta_{r}^{\top} \mathbf{X}_{j}$). $f$ is typically a mapping function, and can be one that applies to dichotomous, ordinal, or continuous distribitions.

>>>>

Scholars working with dyadic data typically begin by stacking observations associated with each dyad on top of one another. This makes sense if each observation is independent of the others. For example, in this design, a conflict initiated from the United States against Japan is assumed to be independent of any conflictual action that Japan may send to the United States. Additionally, every action sent by Japan to others in the system is considered independent even though each of those interactions involves a common sender, i.e, Japan. While most scholars begin with the assumption that each dyadic interaction takes place in isolation of the others, we know this assumption to be false both in theory and in practice. Relational data comes with an explicit structure that generally leads to particular types of dependencies. The importance of accounting for the underlying structure of our data has been a lesson well understood, at least when it comes to time-series cross-sectional data (TSCS) within political science \citep{beck:katz:1995,beck:etal:1998}. As a result, it is now standard practice to take explicit steps to account for the complex data structures that emerge in TSCS applications and the unobserved heterogeneity that they cause.

@@

These types of patterns frequently emerge in IR contexts.\footnote{For example, see: \citet{manger:etal:2012, kinne:2013, chyzh:2016}.} For example, a perennial finding in the interstate trade literature emphasizes the role that geography plays in determining trade flows. Geographic proximity in the network context is an example of homophily---a shared attribute between actors that corresponds to a greater likelihood of the event of interest taking place. Alternatively, in the interstate conflict literature, we may find that actors who are each a member of a particular (formal or informal) alliance are likely to act similarly in the conflict network. Specifically, they will tend to initiate conflictual events with actors that their fellow alliance members initiate conflict with, and they will be unlikely to initiate conflict with members of their alliance---an example of stochastic equivalence. In both these examples, we are able to explicitly parameterize the attribute that might explain the emergence of higher order dependence patterns. While sometimes the conditions driving these patterns, such as geography, are easy to identify, at other times it can be difficult to describe exactly why higher order dependence patterns in networks may develop.

@@

% We use a Bayesian probit framework, in which we model a latent variable, $\theta_{ij}$, first using a set of exogenous dyadic ($\bm\beta_{d}^{\top} \mathbf{X}_{ij}$), sender ($\bm\beta_{s}^{\top} \mathbf{X}_{i}$), and receiver covariates ($\bm\beta_{r}^{\top} \mathbf{X}_{j}$).
where $y_{ij,t}$ captures the interaction between actor $i$ (the sender) and $j$ (the receiver) at time $t$. We model a latent variable, $\theta_{ij}$, first using a set of exogenous dyadic ($\bm\beta_{d}^{\top} \mathbf{X}_{ij}$), sender ($\bm\beta_{s}^{\top} \mathbf{X}_{i}$), and receiver covariates ($\bm\beta_{r}^{\top} \mathbf{X}_{j}$). $f$ is typically a mapping function, and can be one that applies to dichotomous, ordinal, or continuous distributions.

### 2018-10-02 paper/4_app.tex
CD edits

We obtained the data for these studies from their replication archives.\footnote{Without exception the data was easy to retrieve thanks to the authors' transparency and an increasing norm in the social sciences of open data sharing.} The chosen studies have each gained over 100 or more citations and were recently published.\footnote{Note that we chose papers with at least 100 citations as an indicator of influence within the discipline. Of course, by the criteria of influence many other papers could have been considered for the replication.} Each of these pieces, is well-known and posited an hypothesis in which interdependencies are consequential. Reflecting the dominant approach in the literature, the studies tested their hypothesis by employing some form of a generalized linear model. Table~\ref{tab:modelDesign} provides descriptive information for the studies that we replicated.
% \footnote{It is important to note that the AME framework is not well suited for studies where the focus may be on studying onsets as in these type of studies dyad-years representing conflict continuations are removed from the sample. In studies with this type of focus, network models in general are of limited utility since any interdependencies underlying how events may spread from one actor to another are explicitly masked from the estimation process due to the removal of observations.} ........... gibler uses a study onsets. in some ways this is problematic in other it is not, i'm (sm) suggesting that we ignore for now.

@@

The key variables in the original model measure whether or not the sender in the directed dyad is a personalist regime and the target a democratic regime (``Pers/Democ directed dyad") or whether the opposite is true (``Democ/Personalist directed dyad''). The authors find that coefficient of the Pers/Democ Directed Dyad indicator is positive, while the Democ/Personalist directed dyad is too imprecisely measured to indicate a direction. In our re-estimation using the AME framework, we confirm these results, indicating that dictators are more likely to initiate or engage in conflict with democratic regimes but not vice versa.

@@

To test this hypothesis, Gibler employs an undirected dyadic design and estimates his model using a GLM with dyad-clustered standard errors.\footnote{Full tabular results are shown in the appendix, we focus on model 6 from Table 6 (2017, 34).} With this design, Gibler finds support for both the insignificance of contemporary power parity, and the effect of the initial conditions for entry on driving conflict behavior. When we reestimate using AME power parity still has a small, imprecisely measured effect, supporting Gibler's argument that contemporary power parity is an artifact of other aspects of state relations, and does not drive contemporary conflict behavior. At the same time, the AME approach finds that the effect of power parity at entry into the international system becomes too imprecisely measured for using to draw any inferences about its effect on the initiation of MIDs. Not only is our estimate of the effect of this variable small, but it has a very large relative standard error---over a magnitude larger than the parameter itself. Thus, while we can support the claims that, when accounting for both initial conditions and network dependencies, the effect of power parity vanishes, we are able to find less support for the claims about the effects of initial conditions.

@@

First, utilizing the AME framework enables scholars to better model the underlying structure inherent to dyadic data.. In each of the replications, the AME's predictive performance out-of-sample is notably superior. Specifically, not only does AME perform better at correctly identifying cases in which the dependent variable takes a value of $0$ (via the ROC curves), but it also dominates at correctly identifying occurrences of the dependent variable in the data (seen via the PR curves).

>>>>

We obtained the data for these studies from their replication archives.\footnote{Without exception the data was easy to retrieve thanks to the authors' transparency and an increasing norm in the social sciences of open data sharing.} The chosen studies have each gained over 100 or more citations and were recently published.\footnote{Note that we chose papers with at least 100 citations as an indicator of influence within the discipline. Of course, by the criteria of influence many other papers could have been considered for the replication.} Each of these pieces is well-known and posited an hypothesis in which interdependencies are consequential. Reflecting the dominant approach in the literature, the studies tested their hypothesis by employing some form of a generalized linear model. Table~\ref{tab:modelDesign} provides descriptive information for the studies that we replicated.
% \footnote{It is important to note that the AME framework is not well suited for studies where the focus may be on studying onsets as in these type of studies dyad-years representing conflict continuations are removed from the sample. In studies with this type of focus, network models in general are of limited utility since any interdependencies underlying how events may spread from one actor to another are explicitly masked from the estimation process due to the removal of observations.} ........... gibler uses a study onsets. in some ways this is problematic in other it is not, i'm (sm) suggesting that we ignore for now.

@@

The key variables in the original model measure whether or not the sender in the directed dyad is a personalist regime and the target a democratic regime (``Pers/Democ directed dyad") or whether the opposite is true (``Democ/Personalist directed dyad''). The authors find that coefficient of the Pers/Democ Directed Dyad indicator is positive, while the Democ/Personalist directed dyad is too imprecisely measured to indicate a direction. In our re-estimation using the AME framework, we confirm these results, indicating that dictators are likely to initiate or engage in conflict with democratic regimes but not vice versa.

@@

To test this hypothesis, Gibler employs an undirected dyadic design and estimates his model using a GLM with dyad-clustered standard errors.\footnote{Full tabular results are shown in the appendix, we focus on model 6 from Table 6 (2017, 34).} With this design, Gibler finds support for both the insignificance of contemporary power parity, and the effect of the initial conditions for entry on driving conflict behavior. When we reestimate using AME power parity still has a small, imprecisely measured effect, supporting Gibler's argument that contemporary power parity is an artifact of other aspects of state relations, and does not drive contemporary conflict behavior. At the same time, the AME approach finds that the effect of power parity when a country enters into the international system is too imprecisely measured to draw any inferences about its effect on the initiation of MIDs. Not only is our estimate of the effect of this variable small, but it has a very large relative standard error---over a magnitude larger than the parameter itself. Thus, while we can confirm the argument that the effect of power parity vanishes when accounting for both initial conditions and network dependencies, we find less support for the argument that initial conditions drive the initiation of MIDs.

@@

First, utilizing the AME framework enables scholars to better model the underlying structure inherent to dyadic data. In each of the replications, the AME's predictive performance out-of-sample is notably superior. Specifically, not only does AME perform better at correctly identifying cases in which the dependent variable takes a value of $0$ (via the ROC curves), but it also dominates at correctly identifying occurrences of the dependent variable in the data (seen via the PR curves).
